---
title: "Introduction to zeroserve"
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
vignette: >
  %\VignetteIndexEntry{Introduction to zeroserve}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

`zeroserve` provides a high-performance transport layer for serving data from R to web frontends. It is frontend-agnostic and supports various data formats and sources.

## Core Concepts

The package focuses on two main methods of serving data:

1.  **Arrow IPC Streams:** Disk-free, copy-minimized transport for data that fits in R, Mori shared memory, and browser memory. The complete result is serialized before the URL is returned.
2.  **Parquet Files:** Best for larger-than-memory data or data already on disk, served via HTTP with support for range requests.

## Basic Usage

### Serving an Arrow Stream

To serve an `sf` object as an Arrow stream:

```{r}
library(zeroserve)
library(sf)

# Create some sample data
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

# Serve the data
url <- zs_serve_arrow(nc)
print(url)

# Verify the data (e.g., using curl)
# This simulates what a frontend would do
res <- curl::curl_fetch_memory(url)
cat("Fetched", length(res$content), "bytes of Arrow IPC data\n")

# Clean up when done
zs_stop_server()
```

### Serving native DuckSpatial GeoArrow to MapGL

DuckSpatial stays lazy until `zs_serve_arrow()` requests its native GeoArrow stream. ZeroServe then materializes the complete Arrow IPC result in memory and exposes it over HTTP without creating a temporary Arrow payload file.

```{r}
#| eval: false
library(duckspatial)
library(zeroserve)
library(mapgl)
library(deckglgeoarrow)

countries <- ddbs_open_dataset(
  system.file("spatial/countries.geojson", package = "duckspatial")
)

url <- zs_serve_arrow(countries, layer_id = "countries")

maplibre(projection = "mercator") |>
  addSource(id = "countries", url = url) |>
  addGeoArrowPolygonLayer(
    source = "countries",
    layer_id = "countries",
    geom_column_name = attr(countries, "sf_column"),
    tooltip = "NAME_ENGL"
  )
```

This URL route differs from `geoarrowWidget(data =)`, which writes an attachment file. Map renderers expect longitude/latitude coordinates; apply any needed transform lazily in DuckSpatial before calling `zs_serve_arrow()`. For results too large to fit comfortably in memory, filter or aggregate first, or use disk-backed `zs_serve_parquet()`/`zs_serve_file()` when the frontend supports those formats.

### Serving a Parquet File

To serve a DuckDB table as a streamable Parquet file:

```{r}
library(zeroserve)
library(DBI)

# Connect to DuckDB
con <- dbConnect(duckdb::duckdb())

# Load built-in data into DuckDB
dbWriteTable(con, "mtcars", mtcars)

# Serve a query result
url <- zs_serve_parquet(con, "SELECT * FROM mtcars", layer_id = "cars")
print(url)

# Verify the data
# The arrow package can read directly from the URL
df <- arrow::read_parquet(url)
head(df)

# Clean up
zs_stop_server()
dbDisconnect(con)
```

## Advanced: HTTP Range Requests

One of the key features of `zeroserve` is its robust support for HTTP Range requests (RFC 7233). This allows client-side applications (like those using [DuckDB-Wasm](https://duckdb.org/docs/api/wasm/overview)) to fetch only the specific parts of a Parquet file they need.

The handler is hardened to handle various client styles, including those that add extra whitespace or use suffix-byte ranges:

```{r}
# Serve a file
temp_p <- tempfile(fileext = ".parquet")
arrow::write_parquet(mtcars, temp_p)
url <- zs_serve_file(temp_p)

# Fetch a range with non-standard whitespace (handled automatically)
h <- curl::new_handle()
curl::handle_setheaders(h, Range = " bytes = 0 - 99 ")
res <- curl::curl_fetch_memory(url, handle = h)

# Parse and check headers
headers <- curl::parse_headers(res$headers)
print(res$status_code) # 206 (Partial Content)
print(grep("Content-Range", headers, value = TRUE, ignore.case = TRUE))

# Fetch only the last 500 bytes (suffix range)
curl::handle_setheaders(h, Range = "bytes=-500")
res_suffix <- curl::curl_fetch_memory(url, handle = h)
cat("Fetched last", length(res_suffix$content), "bytes\n")
```

## Monitoring and Maintenance

### Server Status

You can check if the server is running using `zs_server_status()`. For a more thorough check, use `ping = TRUE` to verify that the background event loop is actually responsive:

```{r}
# Basic process check
zs_server_status()

# Deep health check (via IPC ping)
zs_server_status(ping = TRUE)
```

### Logging

If you encounter issues, you can inspect the last few lines of the background server log:

```{r}
zs_server_logs(n = 10)
```

## Web Compatibility (CORS)

`zeroserve` is designed to work seamlessly with modern web browsers. It includes built-in support for **CORS (Cross-Origin Resource Sharing)**, including `OPTIONS` preflight requests. This means you can point tools like [Deck.gl](https://deck.gl/) or [geoarrow-js](https://github.com/geoarrow/geoarrow-js) directly at the `zeroserve` URL without needing an intermediate proxy.

CORS is scoped to the URL you were actually given. Every served URL carries a per-resource capability token as its first path segment, and both the `GET` and the `OPTIONS` preflight succeed only for a URL that resolves to a registered resource: the preflight of a token-less or unknown URL is refused with the same opaque 404 as the request itself, carrying no `Access-Control-*` headers at all. So pass the URL returned by `zs_serve_*()` to the browser verbatim rather than rebuilding it by hand. Served responses also carry `Referrer-Policy: no-referrer`, so the token is not forwarded to third-party origins in a `Referer` header -- but it is still a secret in a URL, so do not publish it.

## Next Steps

For more information, please refer to the function documentation.

```{r}
# Clean up
zs_stop_server()
zs_server_status()
```

## Next Steps

For more information, please refer to the function documentation.
