---
title: "Introduction to zeroserve"
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
vignette: >
  %\VignetteIndexEntry{Introduction to zeroserve}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

`zeroserve` provides a high-performance transport layer for streaming large datasets from R to web frontends. It is designed to be frontend-agnostic and supports various data formats and sources.

## Core Concepts

The package focuses on two main methods of serving data:

1.  **Arrow IPC Streams:** Best for in-memory data that should be shared with minimal overhead using shared memory.
2.  **Parquet Files:** Best for larger-than-memory data or data already on disk, served via HTTP with support for range requests.

## Basic Usage

### Serving an Arrow Stream

To serve an `sf` object as an Arrow stream:

```{r}
library(zeroserve)
library(sf)

# Create some sample data
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

# Serve the data
url <- zs_serve_arrow(nc)
print(url)

# Verify the data (e.g., using curl)
# This simulates what a frontend would do
res <- curl::curl_fetch_memory(url)
cat("Fetched", length(res$content), "bytes of Arrow IPC data\n")

# Clean up when done
zs_stop_server()
```

### Serving a Parquet File

To serve a DuckDB table as a streamable Parquet file:

```{r}
library(zeroserve)
library(DBI)

# Connect to DuckDB
con <- dbConnect(duckdb::duckdb())

# Load built-in data into DuckDB
dbWriteTable(con, "mtcars", mtcars)

# Serve a query result
url <- zs_serve_parquet(con, "SELECT * FROM mtcars", layer_id = "cars")
print(url)

# Verify the data
# The arrow package can read directly from the URL
df <- arrow::read_parquet(url)
head(df)

# Clean up
zs_stop_server()
dbDisconnect(con, shutdown = TRUE)
```

## Advanced: HTTP Range Requests

One of the key features of `zeroserve` is its support for HTTP Range requests. This allows client-side applications (like those using [DuckDB-Wasm](https://duckdb.org/docs/api/wasm/overview)) to fetch only the specific parts of a Parquet file they need, rather than downloading the entire file.

```{r}
library(zeroserve)

# Serve a file (e.g., the built-in mtcars as a temp file)
temp_p <- tempfile(fileext = ".parquet")
arrow::write_parquet(mtcars, temp_p)
url <- zs_serve_file(temp_p)

# Fetch only the first 100 bytes using a Range header
h <- curl::new_handle()
curl::handle_setheaders(h, Range = "bytes=0-99")
res <- curl::curl_fetch_memory(url, handle = h)

# Check the response
print(res$status_code) # Should be 206 (Partial Content)
cat("Fetched", length(res$content), "bytes\n")
print(res$headers["content-range"])

# Clean up
zs_stop_server()
```

## Next Steps

For more information, please refer to the function documentation.
